Step-by-Step: Training IMU-based Gestures with Live Feedback

### 2024-02-22 docs/latex-nime/nime-template.tex
minor paper edits - flipped .gif

% Package that enables the use of accents and non 
% standard characters
\usepackage[utf8]{inputenc}

@@

\maketitle
\begin{abstract}
This document serves as an application to NIME's 2024 Student Consortium. Within it, I present my interest in joining the NIME research community, as well as my preliminary results for \textit{Step-by-Step}, a software tool that allows users to train gesture recognition models with live audiovisual feedback. \textit{Step-by-Step} uses a simple neural network to learn to recognize and distinguish multiple gestures in inertial measurement unit (IMU) timeseries data. Users train the model by repeatedly performing a target gesture while receiving live feedback on the model’s ability to recognize that gesture. Once satisfactory results are achieved, the user can save the model and deploy it for realtime musical applications, as demonstrated in supplementary live performance videos. An anonymized code repository is available at \url{https://anonymous.4open.science/r/step-by-step-7506/}
\end{abstract} 
\keywords{Gesture Recognition, IMU, Real-time, Neural Network, Training, Inference, Live Feedback}
\ccsdesc[300]{Applied computing~Sound and music computing}
\ccsdesc[300]{Human-centered computing~Gestural input}
\ccsdesc[300]{Machine learning~Neural networks}

@@

\section{Expected Outcomes}

@@

A common approach to this challenge are \textit{learned} mappings, for which the musician records themselves demonstrating a gesture, then trains a machine learning model to recognize and distinguish the target gesture from other body movements. Indeed, the NIME community has produced several software tools that facilitate this process \cite{bullock:mllib,fiebrink:wekinator, gillian:toolkit, caramiaux:nimeml, mcpherson:pipeline}.

These interactive machine learning tools are typically designed for offline training, wherein the musician records training data, trains the model, then tests it. With no knowledge of how the model will perform until after training, the musician must iterate, repeatedly recording, training, and testing until a satisfactory result is achieved. This opaque, iterative process can be a hindrance to creative flow, especially in live performance settings where improvising novel gesture-sound mappings might be desirable.

Ideally, one would be able to train a model in realtime, receiving live feedback on the model's performance as it learns each time a gesture is performed. Receving model performance reports on a per-gesture basis, the musician can build intuition for what the model has learned (e.g. which variations on the gesture are recognized and whicvh are not), adjust their gesture demonstrations accordingly, and arrive at a satisfactory result more efficiently.

@@

\section{Background}
\subsection{Gesture Recognition}
Gesture recognition is the process of recognizing and distinguishing gestures in data. Gestures are body movement patterns that convey information. In the context of digital musical instruments, gestures are often used to control sound. For example, a musician might use a gesture to control the pitch of a note, the volume of a sound, or the timbre of an instrument.

\subsection{Inertial Measurement Units}
Inertial measurement units (IMUs) are one of many types types of sensors that can be used to measure body movement. IMUs typically include an accelerometer, gyroscope, and magnetometer, each of which measures in three axes: x, y, and z. IMUs are commonly used in wearable devices to track body movements over time. In this work, we use IMUs to measure body movement and recognize gestures.

\subsection{Neural Networks}
Neural networks are one of many classes of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures. In \textit{supervised learning}, neural networks are trained by providing them with examples of inputs and corresponding correct output labels. During training, the model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs. This process is called gradient descent and is performed iteratively until the model converges on a solution. Once trained, the model can be used to make predictions on novel inputs. Here, we perform gradient descent in parallel to the recording and automated labelling of training data, allowing the model to learn in real-time as new examples are provided.

\section{Method}
The system consists of hardware and software components. The hardware measures body movement and streams the data to a computer. The software processes the data, trains a neural network, and provides live feedback to the user. The system is designed to be accessible to users with no machine learning experience while providing a powerful codebase for advanced users.

@@

\subsubsection{Labelling}
The data is labelled by associating each frame with a gesture. A gesture is a sequence of frames that represents a single movement of the body. Here, we focus on percussive gestures and assume that a gesture is complete at the local peak in linear acceleration's magnitude. This constraint allow us to achieve realtime, automated labelling of gestures using a simple peak-detection algorithm. When a local peak meet's user-defined criteria (e.g. acceleration is above a certain threshold), the peak frame is labelled as containing a gesture. All other frames are labelled as containing no gesture.

\begin{figure*}[h]
    \centering
    \includegraphics[width=1.0\textwidth]{../figures/realtime_training.jpg}
    \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance. Bottom Panel: 10 seconds of training data. Acceleration components (x, y, z) are shown in color; magnitude is shown in grey. Each magnitude peak is marked by a thin vertical line, indicating it has been labelled as a gesture in the training data. Top Panel: 10 seconds of inference data. At t = 0 seconds the neural network is completely untrained. Naive to what is and is not a gesture, it predicts gesture and non-gesture with equal probability (0.5). As the model trains on non-gesture date, it begins to predict non-gesture with higher probability. At t = 5 seconds, the model has converged on a solution that reliably distinguishes non-gesture. At t = 10 seconds, the model has converged on a solution that reliably distinguishes gestures. Given more training data and time to learn, the model's performance would continue to improve with gessture and non-gesture probabilities approaching 1 and 0, respectively. See supplementary video for live demo in the github repo.}
    \label{fig:realtime_training}
\end{figure*}

@@

The sensor branch consists of an input layer (short timeseries of recent sensor values) followed by a  a 1D convolutional layer (recognizes patterns in the timeseries). The memory branch consists of an input layer (short timeseries of recent model outputs) followed by a max pooling layer (summarizes the timeseries of recent model outputs based on its highest values). These branches are then concatenated and followed by a fully connected output layer. 
% The model structure is shown in Figure \ref{fig:model_structure}.

@@

In principle this training could be done stochastically, training on each frame's window as it is collected. However, this would  would require a more sophisticated training algorithm that Tensorflow's built-in gradient descent optimizer (at least given the computational performance of the author's machine). For this reason, this proof-of-concept codebase defaults to training on batches of recent windows. When training completes, the model is updated, the training data is updated with recent frames, and training begins again.

@@

\section{Conclusion}
Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data. It is worth noting that the strategies employed here are not limited to IMU data and could be applied to data from other sensors. The authors hope that this work will inspire others to explore the possibilities of realtime machine learning in human-computer interaction.

>>>>

% Package that enables the use of accents and non 
% standard characters
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

@@

\maketitle
\begin{abstract}
This document serves as an application to NIME's 2024 Student Consortium. Within it, I present my interest in joining the NIME research community, as well as my preliminary results for \textit{Step-by-Step}, a software tool that allows users to train gesture recognition models with live audiovisual feedback. \textit{Step-by-Step} uses a simple neural network to learn to recognize and distinguish multiple gestures in inertial measurement unit (IMU) timeseries data. Users train the model by repeatedly performing a target gesture while receiving live feedback on the model’s ability to recognize that gesture. Once satisfactory results are achieved, the user can save the model and deploy it for realtime musical applications. The training process is demonstrated in a supplementary video and an anonymized code repository is available at: \href{https://anonymous.4open.science/r/step-by-step-61D8/}{\textbf{https://anonymous.4open.science/r/step-by-step-61D8/}}

\end{abstract} 
\keywords{Gesture Recognition, IMU, Real-time, Neural Network, Training, Inference, Live Feedback}
\ccsdesc[300]{Applied computing~Sound and music computing}
\ccsdesc[300]{Human-centered computing~Gestural input}
\ccsdesc[300]{Machine learning~Neural networks}

@@

@@

A common approach to this challenge are \textit{learned} mappings, for which the musician records themselves demonstrating a gesture, then trains a machine learning model to recognize and distinguish the target gesture from other body movements. Indeed, the NIME community has prod-\\\\-uced a number of software tools that facilitate this process \cite{bullock:mllib,fiebrink:wekinator, gillian:toolkit, caramiaux:nimeml, mcpherson:pipeline}.

These interactive machine learning tools are typically designed for offline training, wherein the musician records training data, trains the model, then tests it. With no knowledge of how the model will perform until after training, the musician must iterate, repeatedly recording, training, and testing until a satisfactory result is achieved. This iterative process with delayed feedback can be a hindrance to creative flow, especially in live performance settings where improvising novel gesture-sound mappings might be desirable.

Ideally, one would be able to train a model in realtime, receiving live feedback on the model's performance as it learns each time a gesture is performed. Receving model performance reports on a per-gesture basis, the musician can build intuition for what the model has learned (e.g. which variations on the gesture are recognized and which are not), adjust their gesture demonstrations accordingly, and arrive at a satisfactory result more efficiently.

@@

\section{Background}
\subsection{Gesture Recognition}
Gestures are body movement patterns that convey information. Gesture recognition is the process of recognizing and distinguishing gestures in data. In the context of digital musical instruments, gestures are often used to control sound. For example, a musician might use a gesture to control the pitch of a note, the volume of a sound, or the timbre of an instrument.

\subsection{Inertial Measurement Units}
Inertial measurement units (IMUs) are one of many types types of sensors that can be used to measure body movement. IMUs typically include an accelerometer, gyroscope, and magnetometer, each of which measures in three axes: x, y, and z. IMUs are commonly used in wearable devices to track body movements over time. In this work, we use IMUs to measure body movement and recognize gestures.\\\\

\subsection{Neural Networks}
Neural networks are one of many classes of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures. In \textit{supervised learning}, neural networks are trained on example inputs paired with corresponding outputs. During training, the model adjusts its parameters to minimize the difference between its predictions about what the outputs should be and the groundtruth outputs it was provided with. This process is called gradient descent and is performed iteratively until the model converges on a solution. Once trained, the model can be used to make predictions on novel inputs. Here, we perform gradient descent in parallel to the recording and automated labelling of training data, allowing the model to learn in real-time as new examples are provided.

\section{Method}
The system consists of hardware and software components. The hardware measures body movement and streams the data to a computer. The software processes the data, trains a neural network, and provides live feedback to the user.

@@

\subsubsection{Labelling}
The data is labelled by associating each frame with a gesture. A gesture is a sequence of frames that represents a single movement of the body. Here, we focus on percussive gestures and assume that a gesture is complete at the local peak in linear acceleration's magnitude. This constraint allow us to achieve realtime, automated labelling of gestures using a simple peak-detection algorithm. When a local peak meet's user-defined criteria (e.g. acceleration is above a certain threshold), the peak frame is labelled as containing the current target gesture. All other frames are labelled as containing no gesture.

\begin{figure*}[h]
    \centering
    \includegraphics[width=0.98\textwidth]{../figures/realtime_training.jpg}
    \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance. Bottom Panel: 10 seconds of training data. Acceleration components (x, y, z) are shown in color; magnitude is shown in grey. Each magnitude peak is marked by a thin vertical line, indicating it has been automatically recognized and labelled as a gesture in the training data. Top Panel: 10 seconds of inference data. At t = 0 seconds the neural network is completely untrained. Naive to what is and is not a gesture, it predicts gesture and non-gesture with equal probability (0.5). As the model trains on non-gesture data, it begins to predict non-gesture with higher probability. At t = 5 seconds, the model has converged on a solution that reliably distinguishes non-gesture. At t = 10 seconds, the model has converged on a solution that reliably distinguishes gestures. Given more training data and time to learn, the model's performance with continue to improve with gessture and non-gesture probabilities approaching 1 and 0 respectively. The model then can be trained to distinguish additional unique gestures (at least 4 in the author's experience). See supplementary videos for demonstrations.}
    \label{fig:realtime_training}
\end{figure*}

@@

The sensor branch consists of an input layer (short timeseries of recent sensor values) followed by a  a 1D convolutional layer (recognizes patterns in the timeseries). The memory branch consists of an input layer (short timeseries of recent model output values) followed by a max pooling layer (summarizes the timeseries of recent model outputs based on its highest values). These branches are then concatenated and followed by a fully connected output layer. Model architecture is shown in Figure \ref{fig:neural_network}.

Depending on the context, the memory branch may not be necessary. In the context of percussive gestures, the authors have found that this self-reference helps the model to learn that gestures typically do not occur in rapid succession.
% The model structure is shown in Figure \ref{fig:neural_network}.

@@

In principle this training could be done stochastically, training on each frame's window as it is collected. However, this would require a more sophisticated training algorithm than Tensorflow's built-in gradient descent optimizer (at least given the computational performance of the author's machine). For this reason, this proof-of-concept codebase defaults to training on batches of recent windows. When training completes, the model is updated, the training data is updated with recent frames, and training begins again.

@@

\section{Conclusion}
Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data. While currently limitted to percussive gestures (MIDI note on/off), a similar scheme could be used to do realtime learning of continuous mappings (MIDI CC messages) as well. It is worth noting that the strategies employed here are not limited to IMU data and could be applied to data from other sensors.

\section{Expected Outcomes}
In applying for the NIME 2024 Student Consortium, I hope to demo \textit{Step-by-Step} as a preliminary work, receive feedback from the NIME community, connect with other researchers, and experience NIME gesture-sound mapping demonstrations in-person (there is only so much one can glean from reading a paper). Given that NIME-related research is uncommon at my institution, the opportunity to connect with other researchers in the NIME community is particularly exciting. During background research for this submission, I discovered many NIME papers that are relevant to this work, and I am eager to connect with the authors of those papers and others who are interested in similar topics.

\section{Supplementary Videos}
\textit{Please find supplementary videos in the supplementary materials section of the conference application.}
